What can we learn about life in the city by mining the Melbourne’s Open Data Platform?

### 2017-05-05 thesis/proposal_draft.Rmd
update friday - still to do: more detail in research plan, more detail in limitations, more detail in methodology of visualisation and app development.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)
library(ggmap)
library(knitr)
library(lubridate)
```

@@

--------------------------------------------------------------------------------------------------------------------------------------

@@

```{r, out.width= '70%', echo = FALSE, fig.align= "center"}
include_graphics("../img/ped_vis_melb_2.PNG")
```

@@

The purpose of this estimation is purely for the imputation of missing values, so prediction within this period is the highest priority at this point. While it would be desirable, this model is not intended to forecast future values.

Another decision which needs to be made is the starting date to model from. With multiple sensors being installed years later after the initial 28 sensors were installed and capturing data, we do not have a complete set of observations over the total observation period at each sensor location. These locations will need to have pedestrian activity modelled based on neighbouring sensors to predict missing values for the time period prior to the sensor's installation and operation.

@@

The use of weather data to improve forecasting performance would be ideal, due to the availability of weather forecasts compared to many other forms of data. Machine learning techniques will also be explored. In particular, clustering of sensor locations may be useful in identifying patterns at certain types of location, such as tourist attractions and shopping districts.  To evaluate the performance of the model, we look to minimise the RMSE (root mean square error):
$$ \text{RMSE} = \sqrt{\frac{\sum_{t=1}^{T}(\hat{y_t} - y_t)^2}{T}}$$

@@

Once a model for forecasting the data has been established, an interactive web application to present the model will be developed. A key aspect which we want to implement is the ability for the end user to forecast pedestrian activity in the future. Improving on the currently available visualisations, the aesthetics of the visualisation will be designed with human perception theory taken into consideration. Preattentive processing theory will be considered to ensure certain elements of the visualisation allow for quick, preattentive identification of busy and quiet areas. Change blindness theory should also be reviewed in order to avoid misperception of relative changes in pedestrian activity.

For testing the proof of concept, a simple web application for viewing average hourly counts for a given month (for a given year) on a map was developed. 
```{r, fig.align = 'center', echo = FALSE, out.width = "66%"}
include_graphics("../img/ped_vis_melb_me.png")
```

### Limitations

@@

#### Visualisation

### Research Plan
- Imputation that incorporates time and space
- Predicitive model for each sensor
- Visual methods for space-time counts
- Shiny app - interface design, computation
- dynamic

## References  
  1. City of Melbourne, _'About Melbourne Data'_, https://data.melbourne.vic.gov.au/about,  
  _accessed 18 April 2017_  
  2. City of Melbourne, _'Pedestrian Counting System'_,  
  http://www.pedestrian.melbourne.vic.gov.au/pdf/CityofMelbournePedestrianCountingSystem.pdf 
  4 May 2015, _accessed 18 April 2017_  
  3. Healey, Christopher G.,  _'Attention and Visual Memory in Visualization and Computer Graphics'_, IEEE Transactions on Visualization and Computer Graphics,  
  https://www.csc2.ncsu.edu/faculty/healey/download/tvcg.12a.pdf

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 100)
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)
library(ggmap)
library(knitr)
library(lubridate)
library(reshape2)

```

@@

---------------------------------------------------------------------------------------

@@

```{r, out.width= '60%', echo = FALSE, fig.align= "center"}
include_graphics("../img/ped_vis_melb_2.PNG")
```

@@

The purpose of this estimation is purely for the imputation of missing values, so prediction within this period is the highest priority at this point. While it would be desirable, this model is not intended to forecast future values. Evaluating the model's goodness of fit, we perform cross validation. Splitting the data randomly 60/40 into a training set (data for model estimation) and a test set (data set to test the goodness of fit), we can get a quick evaluation of the model's performance. LOOCV (leave one out cross validation) is not computationally feasible due to the large number of observations.

Another decision which needs to be made is the starting date to model from. With multiple sensors being installed years later after the initial 28 sensors were installed and capturing data, we do not have a complete set of observations over the total observation period at each sensor location. Consequently, we will only use data from 2014 onwards, as this appears to be the most complete and comprehensive time period where the obeservations provide useful information.

@@

The use of weather data to improve forecasting performance would be ideal, due to the availability of weather forecasts compared to many other forms of data. Machine learning techniques will also be explored. In particular, clustering of sensor locations may be useful in identifying patterns at certain types of location, such as tourist attractions and shopping districts.  Different models for different types of locations may be used, and clustering may give insights as to which type of model to use. For example, the major city loop train stations, Flinders St, Southern Cross and Melbourne Central may have similar patterns. Then, we would expect to be able to use the same parameterisation of the model to estimate. After a model has been established, we need to evaluate the performance of the model. For this, we have decided to minimise the RMSFE (root mean square forecast error):
$$ \text{RMSFE} = \sqrt{\frac{\sum_{t=1}^{T}(\hat{y_t} - y_t)^2}{T}}$$

@@

Once a model for forecasting the data has been established, visual methods for illustrating the model need to be developed.

Improving on the currently available visualisations, the aesthetics of the visualisation will be designed with human perception theory taken into consideration. Preattentive processing theory will be considered to ensure certain elements of the visualisation allow for quick, preattentive identification of busy and quiet areas. Change blindness theory should also be reviewed in order to avoid misperception of relative changes in pedestrian activity.

### Shiny App Development

To implement the visual methods, an interactive web application to present the model will be developed.

For testing the proof of concept, a simple web application for viewing average hourly counts for a given month (for a given year) on a map was developed. A key aspect which we want to implement is the ability for the end user to forecast pedestrian activity in the future.
```{r, fig.align = 'center', echo = FALSE, out.width = "66%"}
include_graphics("../img/ped_vis_melb_me.png")
```

## Limitations

#### Data  
The pedestrian sensor dataset is very limited in its spatial coverage. There is a possibility of data collected from WiFi devices being made available

@@

#### Visualisation  
Ideally, having a dynamic spatial model would allow us to visualise the pedestrian activity in areas not covered by the currently installed sensors. However, due to time constraints, the visualisation will most likely be based on a model of predictions of pedestrian counts at each sensor location.

## Research Plan
The plan to approach this research project is as follows:

- Imputation that incorporates time and space
- Creating a predicitive model for each sensor
- Investiagtion of visual methods for space-time counts
- Development of a Shiny app: interface design, computation and data backend
- Generation of a dynamic spatial model (if time permits)

## References  
  1. City of Melbourne, _'About Melbourne Data'_, https://data.melbourne.vic.gov.au/about,  
   $\quad$ _accessed 18 April 2017_

  2. City of Melbourne, _'Pedestrian Counting System'_,  
   $\quad$ http://www.pedestrian.melbourne.vic.gov.au/pdf/CityofMelbournePedestrianCountingSystem.pdf  
  4 May 2015, _accessed 18 April 2017_

  3. Gould, Phillip G., Koehler, Anne B., Ord, J Keith., Snyder, Ralph D., Hyndman, Rob J.,  
   $\quad$ Vahid-Araghi, Farshid, (2008) _'Forecasting Time Series with Multiple Seasonal Patterns'_, European Journal of Operational Research 191(1), 207–220

  4. Healey, Christopher G., (2012) _'Attention and Visual Memory in Visualization and Computer Graphics'_, IEEE Transactions on Visualization and Computer Graphics,  
  https://www.csc2.ncsu.edu/faculty/healey/download/tvcg.12a.pdf
\newpage

# Appendix

### Plots of time series at each sensor location over entire time span from 2009:  
```{r, echo = FALSE, cache=TRUE}
ped_plot <- aggregate(x = ped_df$Hourly_Counts, by = list("Sensor_ID" = ped_df$Sensor_ID, "Year" = ped_df$Year, "Month" = ped_df$Month), FUN = sum)

ped_plot$counts <- ped_plot$x

ped_plot$Time <- parse_date(paste(ped_plot$MDate, ped_plot$Month, ped_plot$Year, sep = "-"), format = "%D-%B-%Y")

ped_plot$ID <- as.integer(ped_plot$Sensor_ID)

ggplot(filter(ped_plot, ID < 10)) + geom_line(aes(x = Time, y = counts)) + facet_wrap(~ Sensor_ID)

ggplot(filter(ped_plot, ID > 10, ID < 20)) + geom_line(aes(x = Time, y = counts)) + facet_wrap(~ Sensor_ID)

ggplot(filter(ped_plot, ID > 20, ID < 30)) + geom_line(aes(x = Time, y = counts)) + facet_wrap(~ Sensor_ID)

ggplot(filter(ped_plot, ID > 30, ID < 40)) + geom_line(aes(x = Time, y = counts)) + facet_wrap(~ Sensor_ID)

ggplot(filter(ped_plot, ID > 40)) + geom_line(aes(x = Time, y = counts)) + facet_wrap(~ Sensor_ID)

```

Looking at the plots of the monthly aggregates, we can inspect the data quickly for obvious outliers. After detrending the time series for both the deterministic trends (non-zero mean) and stochastic trends (weekly and daily cycles), we expect to have stationary series.

Note that the Sensor ID #41 is missing from the data for unknown reasons.
